---
title: "Benchmarking DuckDBGRanges"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r BiocStyle::doc_date()`; Modified: 6 July 2026"
package: DuckDBGRanges
vignette: >
  %\VignetteIndexEntry{2. Benchmarking DuckDBGRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      error = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

The *Introduction to DuckDBGRanges* vignette shows how a `DuckDBGRanges` behaves
like an ordinary `GRanges` while keeping its data on disk. This vignette asks the
follow-up question: **what does that cost, and what does it buy?** We compare the
DuckDB backend against an in-memory `r Biocpkg("GenomicRanges")` object on two
range-analysis scenarios at scale:

1. **scATAC-seq peaks**: a million peak-like intervals a few hundred bases wide,
   the shape of multi-sample consensus peak sets;
2. **variants**: ten million SNV-like point positions, the shape of a large
   variant catalogue.

The headline results below were produced **offline** (see [Benchmark setup]) and
are rendered here from a bundled results file, so this vignette builds quickly. The
[A small, live comparison] section runs a miniature version at build time so you
can see the mechanics end to end.

The short version: **DuckDBGRanges wins decisively on memory and on the filtering
that dominates the early stages of a workflow, while in-memory `GRanges` keeps its
edge on the interval-tree operations**, `findOverlaps()`, `subsetByOverlaps()`,
`nearest()`. The two are complementary, which is why the idiomatic pattern is to
*filter with DuckDB, then materialize to `GRanges`* for overlap work.

# What DuckDBGRanges optimizes

`r Biocpkg("DuckDBGRanges")` turns coordinate and metadata filters into SQL that
DuckDB pushes into the Parquet reader, so a region restriction reads only the
matching row groups instead of scanning every range. Existing Bioconductor code
picks this up automatically: calling `restrict()` on a `DuckDBGRanges` *is* a
`WHERE` clause, and chained filters compose into a single lazy query.

| Category | Operations | Backend that wins |
|----------|------------|-------------------|
| Memory footprint | the R object itself | **DuckDBGRanges** (constant ~175 KB) |
| Coordinate / metadata filtering | `restrict`, logical `[`, pipelines | **DuckDBGRanges** (predicate pushdown) |
| Nearest-feature distance | `distanceToNearest` | **DuckDBGRanges** (SQL aggregation) |
| Inter-range transforms | `reduce`, `disjoin` | comparable; DuckDBGRanges pulls ahead at scale |
| Overlap enumeration | `findOverlaps`, `subsetByOverlaps` | GRanges (interval trees) |

The design vignette (*Design and extension of DuckDBGRanges*) shows the SQL each
operation translates to.

# A small, live comparison

To show the mechanics without a large dataset, we build a small range set and run
one operation on both an in-memory `GRanges` and a `DuckDBGRanges`, confirming the
results agree.

```{r live-demo}
library(DuckDBGRanges)
library(GenomicRanges)
library(arrow)

set.seed(1L)
n <- 2000L
df <- data.frame(
    seqnames = sample(paste0("chr", 1:5), n, replace = TRUE),
    start = sample(1:1e6, n),
    strand = sample(c("+", "-", "*"), n, replace = TRUE),
    score = runif(n))
df$end <- df$start + sample(100:500, n, replace = TRUE)

gr <- makeGRangesFromDataFrame(df, keep.extra.columns = TRUE)
path <- tempfile(fileext = ".parquet"); write_parquet(df, path)
ddb <- DuckDBGRanges(path, seqnames = "seqnames", start = "start",
                     end = "end", strand = "strand", mcols = "score")

## same answer, one in memory and one queried from disk
system.time(r_gr  <- restrict(gr,  start = 1L, end = 500000L))
system.time(r_ddb <- restrict(ddb, start = 1L, end = 500000L))
length(r_gr) == length(r_ddb)
```

At this size the in-memory object is faster, there is nothing to gain from going
to disk, and the query has fixed overhead. The advantage appears at scale, which is
what the offline benchmark measures.

# Benchmark setup

The full benchmark generates synthetic ranges at scale across the human autosomes:
**1,000,000** peaks for the scATAC-seq scenario and **10,000,000** SNVs for the
variant scenario, plus a 50,000-feature annotation set used as the subject of the
overlap operations. For each scenario it times six operations on both backends and
records the in-memory footprint of each object.

The backends parallelize differently, so we give each its best effort on the same
core budget: **`GRanges` runs single-threaded C code** (the in-memory baseline),
while **`DuckDBGRanges` autotunes** DuckDB's internal threads up to the budget with
no configuration. The exact parameters are recorded with the results and shown
beneath the tables.

# Results

The tables below report the offline benchmark's measured timings, bundled with the
package as `inst/scripts/benchmark_results.rds`. Regenerate them on your own
hardware with `inst/scripts/run_vignette_benchmarks.R` (see that script's header).

```{r results, echo=FALSE, results='asis'}
helper <- system.file("scripts", "make_timings_table.R", package = "DuckDBGRanges")
if (nzchar(helper)) {
    source(helper)
    res <- load_vignette_timings()
} else {
    res <- NULL
}
if (is.null(res)) {
    cat("_Precomputed benchmark results are not available in this build; ",
        "generate them with `inst/scripts/run_vignette_benchmarks.R`._\n", sep = "")
} else {
    cat("\n**Memory footprint**\n\n")
    print(make_memory_table(
        caption = "In-memory object size; DuckDBGRanges is constant regardless of N.",
        results = res))
    cat("\n\n**scATAC-seq scenario (1M peaks)**\n\n")
    print(make_timings_table("scATAC", results = res,
        caption = "Elapsed seconds. Speedup = GRanges / DuckDBGRanges (>1 favors DuckDB)."))
    cat("\n\n**Variant scenario (10M variants)**\n\n")
    print(make_timings_table("variant", results = res,
        caption = "Elapsed seconds. Speedup = GRanges / DuckDBGRanges (>1 favors DuckDB)."))
    cat("\n\n")
    timings_config_note(res)
}
```

# Summary

The two backends have complementary strengths. A `DuckDBGRanges` is a small,
constant-size handle (about 175 KB) regardless of how many ranges it fronts,
whereas a `GRanges` grows with the data (roughly 150 times smaller for a million
peaks, and over 1,500 times smaller for ten million variants). That is what makes
it practical to reference a very large interval catalogue and pay only for the
queries you actually run.

Operations that reduce data favor the DuckDB backend. `restrict()` and logical
subsetting compile to `WHERE` clauses that read only matching rows, and the
advantage grows with size (a few times faster at a million peaks, about 13 times
at ten million variants). `distanceToNearest()` runs as an SQL aggregation rather
than an interval-tree walk, and shows the largest single speedup measured (about
10 times at a million peaks, about 70 times at ten million variants). The
inter-range transforms `reduce()` and `disjoin()` are closer: in-memory `GRanges`
edges out `reduce()` at a million peaks, but DuckDB pulls ahead on both by ten
million ranges.

Overlap enumeration favors `GRanges`. `findOverlaps()` and `subsetByOverlaps()`
rely on in-memory interval trees that a set-based SQL join does not match
(`GRanges` is roughly 100 times faster), so those are best run on a materialized
`GRanges`. In practice the two compose well: use DuckDB to filter a large set down
cheaply, then hand the small result to `GRanges` for the interval-tree work.

# Backend comparison

| Feature | GRanges | DuckDBGRanges |
|---------|---------|---------------|
| Storage | Memory | Parquet file |
| Object footprint | Proportional to N | Constant (~175 KB) |
| Filtering | Scans all ranges | Predicate pushdown (faster) |
| Nearest-feature distance | Interval trees | SQL aggregation (faster) |
| Overlap enumeration | Interval trees (faster) | SQL join |
| `reduce` / `disjoin` | Optimized C | SQL (comparable; faster at scale) |
| Persistence | Session only | Files persist |
| Read by other languages | R only | R, Python, Julia, ... |

## When to use each

- **`GRanges`**, data fits comfortably in RAM; the fastest choice for
  overlap-heavy work and small-to-mid range sets.
- **`DuckDBGRanges`**, larger-than-memory interval sets, filter-heavy pipelines,
  keeping a huge annotation referenceable at constant memory cost, or when the
  Parquet files are shared with non-R tooling. Combine with `GRanges` via *filter,
  then materialize*.

# Running your own benchmarks

`inst/scripts/run_vignette_benchmarks.R` reproduces these numbers on your hardware
and scales to other range counts (`BENCH_SCATAC_PEAKS`, `BENCH_VARIANTS`,
`BENCH_CORES`; set small counts to smoke-test first). It writes
`benchmark_results.rds`, which this vignette renders via
`inst/scripts/make_timings_table.R`.

For higher-level genomic workflows built on this backend, see the
`r Biocpkg("BiocDuckDB")` package.

# Session information

```{r sessioninfo}
sessionInfo()
```
